---
title: "Reproducible Research: Peer Assessment 1"
author: "Toby Patterson"
date: "15 October 2014"
output:  html_document
keep_md: true
---

```{r setoptions,echo=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Loading and preprocessing the data

We are assuming that the activity.zip file is in the current directory, so let's extract the data into a **activity.csv** file and load it.


```{r loading}
activity_base = read.csv(file='activity.csv',sep=',',header=TRUE,colClasses=c('integer','Date','integer'))
```

## What is mean total number of steps taken per day?

Next let's look at the data.  We're asked to calculate the mean and median, and to make a histogram of the total number of steps taken per day.

```{r basic_exploration}
activity_base_step_mean = mean(activity_base$steps, na.rm=TRUE)
activity_base_step_median = median(activity_base$steps, na.rm=TRUE)
```

The mean of the distribution is `r activity_base_step_mean`, and the median is a statistically unlikely yet expected `r activity_base_step_median`.

|    | Mean   | Median  |
|---:|:------:|:-------:|
|  Base Data | `r activity_base_step_mean`  | `r activity_base_step_median`  |

Note let's look at the histogram.

```{r basic_exploration_histogram}
ggplot(activity_base, aes(date, steps),type='l') + geom_bar(stat = "identity")+ labs(title = "Total Number of Steps by Day", x = "Day", y = "Total Number of Steps")
```

Note that this output warns us that there is missing data.

## What is the average daily activity pattern?

Now aggregate the number of steps by the interval, and plot the results in a series.

```{r daily_activity_plot}
activity_by_day <- with(activity_base, aggregate(steps ~ interval, FUN=mean, na.rm=T))
ggplot(activity_by_day, aes(interval, steps)) + geom_line() + xlab("5 Minute Intervals") + ylab("Number of Average Steps in 5 Minutes") + ggtitle("Frequency of Steps in 5 Minute Intervals")
```

From looking at the chart, it appears that our maximum value is around 205, so lets find out.

```{r daily_activity_max}
activity_by_day_mean = activity_by_day[,1] * activity_by_day[,2] / activity_by_day[,1]
activity_by_day_most = activity_by_day[which.max(activity_by_day_mean),]
```

It looks like our most active 5-minute interval was ```r activity_by_day_most$interval```, with an mean value of ```r activity_by_day_most$steps``` steps in a typical 5-minute interval, or ```r activity_by_day_most$steps / 5``` steps per minute.

## Imputing missing values

Our next task is to find, count, and replace missing values.  We're interested in missing data values, those of the steps.

```{r impute_missing_values_test}
total_missing_values = sum(is.na(activity_base$steps))
```

We have `r total_missing_values` missing values, so lets fix them by filling them with their equivalent average.

```{r impute_missing_values_transform}
activity_full = transform(activity_base, 
  steps = ifelse(is.na(steps), 
    ave(steps, interval, FUN = function(x) mean(x, na.rm = TRUE)), 
    steps)
  )
```

Let's see the effect.

```{r impute_missing_values_report}
activity_full_step_mean = mean(activity_full$steps)
activity_full_step_median = median(activity_full$steps)
ggplot(activity_full, aes(date, steps),type='l') + geom_bar(stat = "identity")+ labs(title = "Total Number of Steps by Day", x = "Day", y = "Total Number of Steps")
```

We can see there are notable differencesin our means, median, and histogram between our initial base dataset and the new full dataset.

|    | Mean   | Median  |
|---:|:------:|:-------:|
|  Base Data | `r activity_base_step_mean`  | `r activity_base_step_median`  |
|  Full Data | `r activity_full_step_mean`  | `r activity_full_step_median`  |

## Are there differences in activity patterns between weekdays and weekends?
