---
title: "Reproducible Research: Peer Assessment 1"
author: "Toby Patterson"
date: "15 October 2014"
output:  html_document
keep_md: true
---

```{r setoptions,echo=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Loading and preprocessing the data

We are assuming that the activity.zip file is in the current directory, so let's extract the data into a **activity.csv** file and load it.


```{r loading}
activity_base = read.csv(file='activity.csv',sep=',',header=TRUE,colClasses=c('integer','Date','integer'))
```

## What is mean total number of steps taken per day?

Next let's look at the data.  We're asked to calculate the mean and median, and to make a histogram of the total number of steps taken per day.

```{r basic_exploration}
activity_base_by_day = with(activity_base, aggregate(steps ~ interval, FUN=mean, na.rm=T))
activity_base_step_mean = mean(activity_base_by_day$steps)
activity_base_step_median = median(activity_base_by_day$steps)
```

The mean of the distribution is `r activity_base_step_mean`, and the median is `r activity_base_step_median`.

|    | Mean   | Median  |
|---:|:------:|:-------:|
|  Base Data | `r activity_base_step_mean`  | `r activity_base_step_median`  |

Note let's look at the histogram.

```{r basic_exploration_histogram}
ggplot(activity_base, aes(date, steps),type='l') + geom_bar(stat = "identity")+ labs(title = "Total Number of Steps by Day", x = "Day", y = "Total Number of Steps")
```

Note that this output warns us that there is missing data.

## What is the average daily activity pattern?

Now aggregate the number of steps by the interval, and plot the results in a series.

```{r daily_activity_plot}

ggplot(activity_base_by_day, aes(interval, steps)) + geom_line() + xlab("5 Minute Intervals") + ylab("Number of Average Steps in 5 Minutes") + ggtitle("Frequency of Steps in 5 Minute Intervals")
```

From looking at the chart, it appears that our maximum value is around 205, so lets find out.

```{r daily_activity_max}
activity_base_by_day_mean = activity_base_by_day[,1] * activity_base_by_day[,2] / activity_base_by_day[,1]
activity_base_by_day_most = activity_base_by_day[which.max(activity_base_by_day_mean),]
```

It looks like our most active 5-minute interval was ```r activity_base_by_day_most$interval```, with an mean value of ```r activity_base_by_day_most$steps``` steps in a typical 5-minute interval, or ```r activity_base_by_day_most$steps / 5``` steps per minute.

## Imputing missing values

Our next task is to find, count, and replace missing values.  We're interested in missing data values, those of the steps.

```{r impute_missing_values_test}
total_missing_values = sum(is.na(activity_base$steps))
```

We have `r total_missing_values` missing values, so lets fix them by filling them with their equivalent average.

```{r impute_missing_values_transform}
activity_full = transform(activity_base, 
  steps = ifelse(is.na(steps), 
    ave(steps, interval, FUN = function(x) mean(x, na.rm = TRUE)), 
    steps)
  )
```

Let's see the effect.

```{r impute_missing_values_report}
activity_full_by_day <- with(activity_full, aggregate(steps ~ interval, FUN=mean, na.rm=T))
activity_full_step_mean = mean(activity_full_by_day$steps)
activity_full_step_median = median(activity_full_by_day$steps)
ggplot(activity_full, aes(date, steps),type='l') + geom_bar(stat = "identity")+ labs(title = "Total Number of Steps by Day", x = "Day", y = "Total Number of Steps")
```

Looking at the new means, median, and histogram, we don't see any major differences in the data.  The only difference appears to be that the histogram now presents data for every day.

|    | Mean   | Median  |
|---:|:------:|:-------:|
|  Base Data | `r activity_base_step_mean`  | `r activity_base_step_median`  |
|  Full Data | `r activity_full_step_mean`  | `r activity_full_step_median`  |

The lack of change in these values between the base and full datasets suggests that our imputing of additional data may not significantly altering our analysis.

## Are there differences in activity patterns between weekdays and weekends?

Now lets compare weekdays to weekends.  To do this, we need to determine if an observation was made on a weekend or weekeday, then plot the data using this new variable as a facet.

```{r detected_differences_by_day}
activity_full$day = factor( ifelse(weekdays(activity_full$date) %in% c("Saturday","Sunday"),'Weekend','Weekday') )
activity_full_by_day <- with(activity_full, aggregate(steps ~ interval + day, FUN=mean, na.rm=T))
ggplot(activity_full_by_day, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) + xlab("5 Minute Intervals") + ylab("Number of Average Steps in 5 Minutes") + ggtitle("Frequency of Steps in 5 Minute Intervals")
```

We can see that the general distribution of intervals retains a similar distribution, with the most prominent peak around 800.  However, our 2nd highest point is not the same: approximately 1600 for weekdays, but either 1250 or 1650 for weekends.